---
title: "Introduction to hvtiRpropensity"
author: "John Ehrlinger"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Introduction to hvtiRpropensity}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hvtiRpropensity)
```

## Overview

`hvtiRpropensity` provides propensity score methods for cardiac surgery
comparative-effectiveness research.  It ports the balancing logic from six SAS
templates into a tidy R API compatible with
[hvtiPlotR](https://ehrlinger.github.io/hvtiPlotR/).

The package covers the full workflow from raw covariates to sensitivity analysis:

1. **Estimate a balancing score** — `ps_logistic()`, `ps_ordinal()`,
   `ps_nominal()`, `bs_continuous()`, `bs_count()`
2. **Balance treatment groups** — `ps_match()` (nearest-neighbour matching)
   or `ps_weight()` (IPTW)
3. **Diagnose balance** — SMD tables, group counts, and overlap diagnostics
   in `$tables`, `sa_overlap()`
4. **Assess sensitivity to unmeasured confounding** — `sa_rosenbaum()`,
   `sa_evalue()`, `sa_overlap()`, `sa_trim_sweep()`

All score, balance, fit, and validation objects share a common `ps_data` S3
base class with four slots: `$data` carries patient-level outputs, `$meta`
records the declared contract and row accounting, `$tables` carries inference
and diagnostics, and `$models` retains every fitted model used to create the
result.

---

## Sample data

`sample_ps_data()` generates a synthetic cardiac-surgery dataset that mimics a
typical SAS export: two treatment arms (SAVR / TF-TAVR), several continuous and
binary covariates, a fitted propensity score (`prob_t`), and a `match` indicator
initialised to `0`.

```{r}
#| label: sample-data
dta <- sample_ps_data(n = 500, seed = 42)
head(dta)
```

```{r}
#| label: sample-data-summary
table(dta$tavr)
summary(dta$prob_t)
```

The two treatment groups differ on all covariates — younger patients with higher
ejection fraction were more likely to receive TF-TAVR — so the raw SMDs are
non-trivial and the propensity scores have good overlap.

```{r}
#| label: ps-hist
hist(
  dta$prob_t,
  breaks = 30,
  main   = "Propensity score distribution",
  xlab   = "Estimated probability of TF-TAVR",
  col    = "steelblue",
  border = "white"
)
```

---

## Outcome-model bundles

`fit_logistic()` fits binary, proportional-odds ordinal, and generalized-logit
nominal outcome models. The complete outcome levels and the event or reference
level are required. This makes the scientific contrast independent of factor
storage order.

```{r}
#| label: fit-logistic-complete
outcome_model <- fit_logistic(
  tavr ~ age + female + ef + diabetes + hypertension,
  data = dta,
  family = "binary",
  outcome_levels = c(0, 1),
  event_level = 1
)
outcome_model$tables$estimates
outcome_model$tables$fit_status
```

Every bundle records `bundle_version = 1L`, the formula, required predictors,
declared outcome levels, package versions, analysis counts, all model objects,
and four common inference tables:

| table | schema |
|---|---|
| `estimates` | `term`, `estimate`, `std.error`, `statistic`, `df`, `p.value`, `conf.low`, `conf.high`, `odds_ratio`, `pooled` |
| `covariance` | named coefficient covariance matrix |
| `by_imputation` | `imputation`, `term`, `estimate`, `std.error` |
| `fit_status` | `imputation`, `converged`, `n_input`, `n_analyzed`, `n_excluded` |

For stacked long-form imputations, the same model is fit independently within
each imputation. Predictions are averaged patient by patient, matching the
legacy `PROC SUMMARY mean=` workflow, while coefficient inference and
covariance use Rubin's rules. These answer different questions and are stored
separately. A missing patient, changed outcome or categorical predictor level,
aliased term, failed fit, or unusable covariance stops pooling.

```{r}
#| label: fit-logistic-mi
dta_mi <- do.call(rbind, lapply(1:3, function(imputation) {
  partition <- dta
  partition$age <- partition$age + (imputation - 2) / 10
  partition$imputation <- imputation
  partition
}))

outcome_model_mi <- fit_logistic(
  tavr ~ age + female + ef + diabetes + hypertension,
  data = dta_mi,
  family = "binary",
  id_col = "id",
  imputation_col = "imputation",
  outcome_levels = c(0, 1),
  event_level = 1
)
outcome_model_mi$tables$estimates
outcome_model_mi$tables$fit_status
```

`validate_logistic()` scores a validation cohort from a saved binary bundle
without refitting and without modifying the source object. It returns
equal-frequency calibration groups and a performance row containing observed
and expected event counts, O:E ratio, AUC, and Brier score. The first interface
is intentionally binary only.

```{r}
#| label: validate-logistic
validation <- validate_logistic(outcome_model, dta, groups = 10L)
validation$tables$performance
```

---

## Propensity score estimation

### Binary treatment — `ps_logistic()`

In practice a propensity score is estimated from covariates before matching or
weighting.  `ps_logistic()` fits `stats::glm(..., family = binomial())` and
appends several columns that mirror the original SAS template output:

| R column | SAS equivalent | Description |
|---|---|---|
| `prob_t` | `_p_` / `_propen_` | Propensity score (probability of treatment) |
| `logit_t` | `_logit_` | Log-odds: `log(p / (1 - p))` |
| `mt_wt` | `mt_wt` | Matching weight `min(p, 1-p) / (p·trt + (1-p)·(1-trt))` |
| `quintile` | `quintile` | Rank-based quintile (1–5) |
| `decile` | `decile` | Rank-based decile (1–10) |

```{r}
#| label: ps-logistic
obj_ps <- ps_logistic(
  tavr ~ age + female + ef + diabetes + hypertension,
  data = sample_ps_data(n = 500, seed = 42)
)
print(obj_ps)
```

The scored dataset can be passed directly to `ps_match()` or `ps_weight()`.

```{r}
#| label: ps-logistic-cols
head(obj_ps$data[, c("id", "tavr", "prob_t", "logit_t", "mt_wt",
                      "quintile", "decile")])
```

```{r}
#| label: ps-logistic-to-match
m_from_ps <- ps_match(obj_ps$data,
                      score_col = obj_ps$meta$score_col,
                      seed      = 42)
print(m_from_ps)
```

### Multiply-imputed data

The SAS template fits the model separately on each imputed dataset
(`PROC LOGISTIC ... BY _IMPUTATION_`) and averages predicted probabilities
with `PROC SUMMARY mean=`.  Pass a stacked long-format MI data frame and set
`imputation_col` to replicate this.

```{r}
#| label: ps-logistic-mi
dta_base <- sample_ps_data(n = 300, seed = 10)

# Simulate two imputed datasets stacked with an imputation index column
dta_mi <- rbind(cbind(dta_base, imp = 1L),
                cbind(dta_base, imp = 2L))

obj_mi <- ps_logistic(
  tavr ~ age + female + ef + diabetes + hypertension,
  data           = dta_mi,
  imputation_col = "imp",
  id_col         = "id"
)
print(obj_mi)

# Per-patient PS is averaged across imputations, matching PROC SUMMARY mean=
summary(obj_mi$data$prob_t)
```

### Ordinal treatment — `ps_ordinal()`

For ordered treatment variables (e.g. NYHA functional class I / II / III–IV),
`ps_ordinal()` fits a proportional-odds cumulative logit model via
`MASS::polr()`, equivalent to the default `PROC LOGISTIC` behaviour for ordinal
responses in SAS.

The SAS template decomposes cumulative probabilities into marginal probabilities:
`p1 = col1; p2 = col2 - col1; p3 = 1 - col2`.  `ps_ordinal()` performs this
decomposition internally, appending one `prob_<level>` column per level.

```{r}
#| label: ps-ordinal
#| eval: !expr requireNamespace("MASS", quietly = TRUE)
dta_ord <- sample_ps_data_ordinal(n = 400, seed = 1)
obj_ord  <- ps_ordinal(
  nyha_grp ~ age + female + ef + diabetes,
  data = dta_ord
)
print(obj_ord)

# One marginal probability column per level
head(obj_ord$data[, c("id", "nyha_grp",
                       obj_ord$meta$score_cols)])

# Quintile distribution (ordered by P(highest level) as in SAS)
table(obj_ord$data$quintile)
```

### Nominal treatment — `ps_nominal()`

For unordered multi-level treatments (e.g. repair type: CE / COS / PER / DEV),
`ps_nominal()` fits a generalised logit model via `nnet::multinom()`, equivalent
to SAS `PROC LOGISTIC ... / link=glogit`.  The first factor level is the
reference category (`REF=first` in SAS); use `ref_level` to override.

```{r}
#| label: ps-nominal
#| eval: !expr requireNamespace("nnet", quietly = TRUE)
dta_nom <- sample_ps_data_nominal(n = 400, seed = 2)
obj_nom  <- ps_nominal(
  rtyp ~ age + female + ef + diabetes,
  data = dta_nom
)
print(obj_nom)

# One probability column per repair type (analogous to p_cos, p_per, etc.)
head(obj_nom$data[, c("id", "rtyp", obj_nom$meta$score_cols)])
```

---

## Balancing scores for continuous and count exposures

When the "treatment" is a continuous or count variable (e.g. units of RBC
transfused, nadir haematocrit), a propensity score is undefined.  The SAS
templates instead estimate a **balancing score**: the linear predictor from a
saturated model.  Patients with similar balancing scores have similar expected
exposure values, enabling within-stratum comparisons analogous to PS
stratification.

### Continuous exposure — `bs_continuous()`

Mirrors `tp.rm.continuous.balncing_score.sas` (PROC REG balancing score on
nadir HCT).  The fitted value from `stats::lm()` is used as the score.

```{r}
#| label: bs-continuous
dta <- sample_ps_data(n = 400, seed = 5)
obj_bs <- bs_continuous(
  ef ~ age + female + diabetes + hypertension,
  data   = dta,
  id_col = "id"
)
print(obj_bs)

# Balancing score and decile-based strata
head(obj_bs$data[, c("id", "ef", "bs", "quintile", "decile", "cluster")])

# Binary stratum indicators stra_1 ... stra_10 mirror the SAS %grp macro
table(obj_bs$data$cluster)
```

### Count exposure — `bs_count()`

Mirrors `tp.pm.count.balncing_score.sas` (PROC GENMOD dist=nb link=log).  The
**linear predictor on the log scale** (`xbeta`) is used as the balancing score,
not the fitted count — the log scale is continuous and unbounded, making it
more suitable for rank-based stratification.

```{r}
#| label: bs-count
#| eval: !expr requireNamespace("MASS", quietly = TRUE)
dta_cnt <- sample_ps_data_count(n = 400, seed = 6)

# Negative binomial to account for overdispersion (SAS dist=nb)
obj_cnt <- bs_count(
  rbc_tot ~ female + age + diabetes + hypertension,
  data = dta_cnt,
  dist = "negbin"
)
print(obj_cnt)

head(obj_cnt$data[, c("id", "rbc_tot", "bs", "cluster",
                       "stra_1", "stra_2")])
```

---

## Propensity score matching

### Build the matched object

`ps_match()` performs greedy 1:1 nearest-neighbour matching without
replacement.

```{r}
#| label: match-build
m <- ps_match(dta)
print(m)
```

### Diagnostics

`summary()` prints the available diagnostic tables.

```{r}
#| label: match-summary
summary(m)
```

The `$tables` slot is a named list you can access directly:

```{r}
#| label: match-tables
m$tables$smd_before
m$tables$smd_after
```

### Extract the matched dataset

`$data` is the *full* dataset with a `match` column set to `1` for matched
pairs.  Filter to `match == 1` to obtain the matched subset.

```{r}
#| label: match-data
matched <- m$data[m$data$match == 1L, ]
nrow(matched)
table(matched$tavr)
```

### Caliper matching

A caliper restricts matches to pairs whose propensity scores differ by no more
than the supplied threshold, reducing the number of matched pairs in exchange
for tighter balance.

```{r}
#| label: match-caliper
m_cal <- ps_match(dta, caliper = 0.05)
print(m_cal)
```

---

## Inverse-probability-of-treatment weighting

### ATE weights (default)

`ps_weight()` computes IPTW weights and appends them to `$data`.  The default
estimand is the **average treatment effect** (ATE).

```{r}
#| label: weight-ate
w_ate <- ps_weight(dta, estimand = "ATE")
print(w_ate)
```

```{r}
#| label: weight-ate-summary
summary(w_ate)
```

The `iptw` column holds the stabilised weights.

```{r}
#| label: weight-ate-head
head(w_ate$data[, c("id", "tavr", "prob_t", "iptw")])
summary(w_ate$data$iptw)
```

### ATT and ATC weights

For the **average treatment effect on the treated** (ATT), treated patients
receive weight `1` and controls receive `ps / (1 - ps)`.

```{r}
#| label: weight-att
w_att <- ps_weight(dta, estimand = "ATT")
summary(w_att$data$iptw)
```

For the **average treatment effect on the controls** (ATC), the roles are
reversed.

```{r}
#| label: weight-atc
w_atc <- ps_weight(dta, estimand = "ATC")
summary(w_atc$data$iptw)
```

### Weight trimming (winsorisation)

Extreme weights can destabilise estimates.  Supplying `trim` winsorises weights
to the (`trim`, `1 - trim`) quantile range.

```{r}
#| label: weight-trim
w_trim <- ps_weight(dta, estimand = "ATE", trim = 0.05)
summary(w_trim$data$iptw)
```

---

## Sensitivity analysis

Every observational study rests on the assumption of no unmeasured confounding.
The four `sa_*` functions quantify how strong an unmeasured confounder would
need to be to change the study's conclusions.

```{r}
#| label: sa-setup
dta  <- sample_ps_data(n = 500, seed = 42)
m    <- ps_match(dta, seed = 42)
w    <- ps_weight(dta, estimand = "ATE")
```

### Overlap and positivity — `sa_overlap()`

Before any analysis, check that the PS distributions overlap sufficiently.
`sa_overlap()` reports the common support region, the fraction of each group
outside it, and near-positivity-violation flags.

```{r}
#| label: sa-overlap
ov <- sa_overlap(m)
ov$overlap_region
ov$outside_overlap
ov$positivity_flags
```

### Weight-trimming sensitivity sweep — `sa_trim_sweep()`

Extreme IPTW weights inflate variance and can dominate estimates.
`sa_trim_sweep()` sweeps a range of winsorisation thresholds and reports how
ESS and the weight distribution change.  Use this to choose a trimming level
that controls extreme weights without unacceptable loss of sample size.

```{r}
#| label: sa-trim-sweep
sweep <- sa_trim_sweep(w, trim_seq = seq(0, 0.10, by = 0.01))
head(sweep)

# ESS at no trimming vs. 5% trimming
sweep[sweep$trim %in% c(0, 0.05), c("trim", "ess_control",
                                      "ess_treated", "max_weight")]
```

### E-values — `sa_evalue()`

The E-value (VanderWeele & Ding 2017) is the minimum risk-ratio that an
unmeasured confounder would need with *both* the treatment and the outcome to
fully explain away the observed association.  A large E-value indicates a more
robust finding.

```{r}
#| label: sa-evalue-rr
# Suppose a matched analysis yields RR = 1.8 (95% CI 1.3 to 2.5)
ev <- sa_evalue(1.8, ci_lo = 1.3, ci_hi = 2.5, type = "RR")
ev$evalue_estimate   # confounder must be this strong to explain the RR
ev$evalue_ci         # must be this strong to explain even the CI lower bound
```

```{r}
#| label: sa-evalue-rd
# Risk difference of 0.08 with baseline outcome probability 0.15
ev_rd <- sa_evalue(0.08, ci_lo = 0.02, ci_hi = 0.14,
                   type = "RD", p0 = 0.15)
ev_rd$evalue_estimate
```

### Rosenbaum sensitivity bounds — `sa_rosenbaum()`

For matched analyses, `sa_rosenbaum()` computes worst- and best-case p-value
bounds for the Wilcoxon signed-rank test under a hidden-bias parameter Γ
(Rosenbaum 2002).  Γ = 1 means no hidden bias; the **sensitivity value** is
the largest Γ at which the analysis remains significant.

```{r}
#| label: sa-rosenbaum
res <- sa_rosenbaum(m, outcome_col = "ef", gamma_max = 3, gamma_inc = 0.25)

# Summary: how sensitive is the result?
res$sensitivity_value   # largest Gamma where upper p-value < 0.05
res$n_pairs

# Full bounds table
head(res$bounds)
```

---

## Downstream use with hvtiPlotR

The `$data` and `$tables` slots are designed for direct consumption by
`hvtiPlotR::hv_mirror_hist()` (weighted mode) and `hvtiPlotR::hv_balance()`.

### Mirror histogram (matched)

```{r}
#| label: mirror-hist-match
#| eval: false
library(hvtiPlotR)

mh <- hv_mirror_hist(
  data          = m$data,
  x             = prob_t,
  treatment     = tavr,
  match_col     = match,
  group_labels  = c("SAVR", "TF-TAVR")
)
plot(mh) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::labs(x = "Propensity Score", y = "Number of Patients") +
  hv_theme("manuscript")
```

### Mirror histogram (weighted)

```{r}
#| label: mirror-hist-weight
#| eval: false
mh_wt <- hv_mirror_hist(
  data          = w_ate$data,
  x             = prob_t,
  treatment     = tavr,
  weight_col    = iptw,
  group_labels  = c("SAVR", "TF-TAVR")
)
plot(mh_wt) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::labs(x = "Propensity Score", y = "Weighted Count") +
  hv_theme("manuscript")
```

---

## Session info

```{r}
#| label: session-info
sessionInfo()
```
